---
title: "HCV amino-acid covariation analysis"
output: html_document
params:
  input_dir: "."
  output_dir: "."
  permutations: 999
  seed: 20260923
  min_n: 20
  min_minor: 3
  alpha: 0.05
  min_combination_count: 3
  include_all_position_scan: false
  near_duplicate_differences: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(stringsAsFactors = FALSE)

ras_positions <- list(
  NS3 = c(36, 41, 43, 54, 55, 56, 80, 122, 155, 156, 158, 166, 168, 170, 175),
  NS5A = c(24, 26, 28, 29, 30, 31, 32, 38, 58, 62, 92, 93),
  NS5B = c(150, 159, 206, 282, 316, 320, 321)
)

if (!dir.exists(params$output_dir)) dir.create(params$output_dir, recursive = TRUE)
```

# Purpose

For each position pair, this report measures amino-acid association **within
subtype**, combines the subtype-specific mutual information (MI) values, and
uses within-subtype permutations to obtain P values. It writes CSV result
tables next to the input files (or to `params$output_dir`).

The report produces two overall summaries for every pair:

- **Sequence-weighted:** subtypes contribute in proportion to their usable
  sequence count.
- **Subtype-balanced:** each eligible subtype contributes equally.

It runs both the complete data and an exact-deduplication sensitivity analysis.
Exact deduplication uses all numeric columns in the input range within subtype
and retains incomplete profiles without deduplicating them.

# Functions

```{r functions}
gene_for <- function(path) {
  matches <- names(ras_positions)[vapply(names(ras_positions), function(gene) {
    grepl(gene, basename(path), fixed = TRUE)
  }, logical(1))]
  if (length(matches) != 1L) stop("Cannot identify NS3, NS5A, or NS5B from: ", path)
  matches
}

read_profile <- function(path) {
  dat <- read.csv(path, check.names = FALSE, stringsAsFactors = FALSE,
                  na.strings = character())
  positions <- names(dat)[grepl("^[0-9]+$", names(dat))]
  required <- c("Genotype", "Subtype")
  if (!all(required %in% names(dat))) {
    stop(basename(path), " is missing: ", paste(setdiff(required, names(dat)), collapse = ", "))
  }
  if (!length(positions)) stop(basename(path), " has no numeric position columns")

  dat[positions] <- lapply(dat[positions], function(x) {
    x <- trimws(as.character(x))
    x[x == ""] <- NA_character_
    x
  })
  dat$Subtype <- trimws(as.character(dat$Subtype))
  if (anyNA(dat$Subtype) || any(dat$Subtype == "")) stop(basename(path), " has an empty Subtype")
  list(dat = dat, positions = positions[order(as.integer(positions))])
}

exact_deduplicate <- function(dat, positions) {
  complete_profile <- complete.cases(dat[, positions, drop = FALSE])
  complete_dat <- dat[complete_profile, , drop = FALSE]
  incomplete_dat <- dat[!complete_profile, , drop = FALSE]
  profile_key <- do.call(paste, c(complete_dat[, positions, drop = FALSE], sep = "\r"))
  key <- paste(complete_dat$Subtype, profile_key, sep = "|")
  rbind(complete_dat[!duplicated(key), , drop = FALSE], incomplete_dat)
}

profile_distance <- function(left, right, maximum) {
  differences <- 0L
  for (i in seq_along(left)) {
    if (left[i] != right[i]) {
      differences <- differences + 1L
      if (differences > maximum) return(differences)
    }
  }
  differences
}

complete_linkage_deduplicate <- function(dat, positions, maximum_differences) {
  if (maximum_differences < 0) stop("near_duplicate_differences must be non-negative")
  complete <- complete.cases(dat[, positions, drop = FALSE])
  incomplete_dat <- dat[!complete, , drop = FALSE]
  complete_dat <- dat[complete, , drop = FALSE]
  retained <- list(incomplete_dat)

  for (subtype in sort(unique(complete_dat$Subtype))) {
    candidates <- complete_dat[complete_dat$Subtype == subtype, , drop = FALSE]
    profiles <- lapply(seq_len(nrow(candidates)), function(i) as.character(candidates[i, positions]))
    clusters <- lapply(seq_len(nrow(candidates)), function(i) i)
    repeat {
      merge <- NULL
      for (left in seq_len(length(clusters) - 1L)) {
        for (right in seq.int(left + 1L, length(clusters))) {
          all_close <- all(vapply(clusters[[left]], function(i) {
            all(vapply(clusters[[right]], function(j) {
              profile_distance(profiles[[i]], profiles[[j]], maximum_differences) <= maximum_differences
            }, logical(1)))
          }, logical(1)))
          if (all_close) {
            merge <- c(left, right)
            break
          }
        }
        if (!is.null(merge)) break
      }
      if (is.null(merge)) break
      clusters[[merge[1]]] <- c(clusters[[merge[1]]], clusters[[merge[2]]])
      clusters[[merge[2]]] <- NULL
    }
    retained[[length(retained) + 1L]] <- candidates[vapply(clusters, `[`, integer(1), 1L), , drop = FALSE]
  }
  do.call(rbind, retained)
}

mutual_information <- function(x, y) {
  tab <- table(x, y)
  if (nrow(tab) < 2L || ncol(tab) < 2L) return(0)
  p <- tab / sum(tab)
  px <- rowSums(p)
  py <- colSums(p)
  expected <- outer(px, py)
  present <- p > 0
  sum(p[present] * log2(p[present] / expected[present]))
}

eligible_subtypes <- function(dat, pos1, pos2, min_n, min_minor) {
  output <- list()
  for (subtype in sort(unique(dat$Subtype))) {
    d <- dat[dat$Subtype == subtype, c(pos1, pos2), drop = FALSE]
    d <- d[complete.cases(d), , drop = FALSE]
    if (nrow(d) < min_n) next
    if (nrow(d) - max(table(d[[pos1]])) < min_minor) next
    if (nrow(d) - max(table(d[[pos2]])) < min_minor) next
    output[[subtype]] <- list(x = d[[pos1]], y = d[[pos2]], n = nrow(d))
  }
  output
}

weighted_mi <- function(mi, sizes, weighting) {
  if (weighting == "sequence_weighted") return(sum(mi * sizes) / sum(sizes))
  if (weighting == "subtype_balanced") return(mean(mi))
  stop("Unknown weighting: ", weighting)
}

analyze_pair <- function(dat, pos1, pos2, permutations, min_n, min_minor) {
  by_subtype <- eligible_subtypes(dat, pos1, pos2, min_n, min_minor)
  sizes <- vapply(by_subtype, function(z) z$n, numeric(1))
  base <- data.frame(
    pos1 = as.integer(pos1), pos2 = as.integer(pos2),
    n_subtypes = length(sizes), n_sequences = sum(sizes)
  )
  weighting <- c("sequence_weighted", "subtype_balanced")
  if (!length(by_subtype)) {
    return(cbind(base[rep(1, 2), ], weighting = weighting,
                 observed_mi_bits = NA_real_, mean_permuted_mi_bits = NA_real_,
                 adjusted_mi_bits = NA_real_, p_value = NA_real_))
  }

  observed_by_subtype <- vapply(by_subtype, function(z) mutual_information(z$x, z$y), numeric(1))
  observed <- vapply(weighting, function(w) weighted_mi(observed_by_subtype, sizes, w), numeric(1))
  permuted <- matrix(NA_real_, nrow = permutations, ncol = length(weighting),
                     dimnames = list(NULL, weighting))
  for (b in seq_len(permutations)) {
    permuted_by_subtype <- vapply(by_subtype, function(z) {
      mutual_information(z$x, sample(z$y, length(z$y), replace = FALSE))
    }, numeric(1))
    permuted[b, ] <- vapply(weighting, function(w) {
      weighted_mi(permuted_by_subtype, sizes, w)
    }, numeric(1))
  }
  data.frame(
    base[rep(1, 2), ], weighting = weighting,
    observed_mi_bits = observed,
    mean_permuted_mi_bits = colMeans(permuted),
    adjusted_mi_bits = observed - colMeans(permuted),
    p_value = vapply(seq_along(weighting), function(i) {
      (1 + sum(permuted[, i] >= observed[i])) / (permutations + 1)
    }, numeric(1))
  )
}

analyze_pairs <- function(dat, pairs, permutations, min_n, min_minor, label) {
  started <- Sys.time()
  results <- vector("list", nrow(pairs))
  for (i in seq_len(nrow(pairs))) {
    results[[i]] <- analyze_pair(dat, pairs[i, 1], pairs[i, 2], permutations, min_n, min_minor)
    if (i %% 100L == 0L || i == nrow(pairs)) {
      elapsed <- as.numeric(difftime(Sys.time(), started, units = "mins"))
      cat(sprintf("    %s: %d/%d pairs (%.1f%%), elapsed %.1f min\n",
                  label, i, nrow(pairs), 100 * i / nrow(pairs), elapsed))
    }
  }
  results <- do.call(rbind, results)
  results$q_value <- NA_real_
  for (weight in unique(results$weighting)) {
    selected <- results$weighting == weight & !is.na(results$p_value)
    results$q_value[selected] <- p.adjust(results$p_value[selected], method = "BH")
  }
  results[order(results$weighting, results$q_value, -results$adjusted_mi_bits, na.last = TRUE), ]
}

subtype_mi_rows <- function(dat, selected_pairs, min_n, min_minor) {
  output <- list()
  index <- 0L
  for (i in seq_len(nrow(selected_pairs))) {
    pos1 <- as.character(selected_pairs[i, 1])
    pos2 <- as.character(selected_pairs[i, 2])
    by_subtype <- eligible_subtypes(dat, pos1, pos2, min_n, min_minor)
    for (subtype in names(by_subtype)) {
      z <- by_subtype[[subtype]]
      index <- index + 1L
      output[[index]] <- data.frame(pos1 = as.integer(pos1), pos2 = as.integer(pos2),
                                    subtype = subtype, n_sequences = z$n,
                                    mi_bits = mutual_information(z$x, z$y))
    }
  }
  if (!length(output)) return(data.frame(pos1 = integer(), pos2 = integer(), subtype = character(),
                                         n_sequences = integer(), mi_bits = numeric()))
  do.call(rbind, output)
}

combination_rows <- function(dat, selected_pairs, min_n, min_minor, minimum_count) {
  output <- list()
  index <- 0L
  for (i in seq_len(nrow(selected_pairs))) {
    pos1 <- as.character(selected_pairs[i, 1])
    pos2 <- as.character(selected_pairs[i, 2])
    by_subtype <- eligible_subtypes(dat, pos1, pos2, min_n, min_minor)
    for (subtype in names(by_subtype)) {
      z <- by_subtype[[subtype]]
      tab <- table(z$x, z$y)
      for (row in seq_len(nrow(tab))) for (column in seq_len(ncol(tab))) {
        observed <- tab[row, column]
        if (observed < minimum_count) next
        expected <- sum(tab[row, ]) * sum(tab[, column]) / z$n
        index <- index + 1L
        output[[index]] <- data.frame(
          pos1 = as.integer(pos1), aa1 = rownames(tab)[row],
          pos2 = as.integer(pos2), aa2 = colnames(tab)[column], subtype = subtype,
          n_sequences = z$n, observed_count = observed,
          expected_count_independent = expected,
          lift_observed_over_expected = observed / expected,
          log2_lift = log2(observed / expected)
        )
      }
    }
  }
  if (!length(output)) return(data.frame(
    pos1 = integer(), aa1 = character(), pos2 = integer(), aa2 = character(),
    subtype = character(), n_sequences = integer(), observed_count = integer(),
    expected_count_independent = numeric(), lift_observed_over_expected = numeric(), log2_lift = numeric()
  ))
  do.call(rbind, output)
}

write_scan_outputs <- function(prefix, family, results, dat) {
  result_path <- file.path(params$output_dir, paste0(prefix, "_", family, ".csv"))
  write.csv(results, result_path, row.names = FALSE)
  selected <- unique(results[!is.na(results$q_value) & results$q_value <= params$alpha, c("pos1", "pos2")])
  detail_prefix <- file.path(params$output_dir, paste0(prefix, "_", family, "_significant"))
  write.csv(subtype_mi_rows(dat, selected, params$min_n, params$min_minor),
            paste0(detail_prefix, "_subtype_mi.csv"), row.names = FALSE)
  write.csv(combination_rows(dat, selected, params$min_n, params$min_minor,
                             params$min_combination_count),
            paste0(detail_prefix, "_aa_combinations.csv"), row.names = FALSE)
  eligible <- length(unique(results[!is.na(results$p_value), c("pos1", "pos2")]))
  cat(sprintf("    complete: %d eligible pairs; %d pairs with q <= %g; wrote %s\n",
              eligible, nrow(selected), params$alpha, basename(result_path)))
}
```

# Run the analyses

```{r analyze}
input_paths <- list.files(params$input_dir, pattern = "_Wide_.*\\.csv$", full.names = TRUE)
input_paths <- input_paths[!grepl("_covariation_", basename(input_paths), fixed = TRUE)]
if (!length(input_paths)) stop("No original *_Wide_*.csv input files found in: ", params$input_dir)

cat(sprintf("Covariation analysis: %d input file(s); %d permutations; min_n=%d; min_minor=%d; seed=%d\n",
            length(input_paths), params$permutations, params$min_n, params$min_minor, params$seed))
set.seed(params$seed)

for (path in input_paths) {
  profile <- read_profile(path)
  dat <- profile$dat
  positions <- profile$positions
  gene <- gene_for(path)
  ras <- ras_positions[[gene]]
  if (!all(as.character(ras) %in% positions)) {
    stop(basename(path), " is missing RAS positions: ",
         paste(setdiff(ras, as.integer(positions)), collapse = ", "))
  }
  non_ras <- setdiff(as.integer(positions), ras)
  families <- list(
    ras_ras = t(combn(ras, 2)),
    ras_nonras = as.matrix(expand.grid(ras, non_ras, stringsAsFactors = FALSE))
  )
  if (isTRUE(params$include_all_position_scan)) {
    families$all_positions <- t(combn(as.integer(positions), 2))
  }

  complete_profiles <- sum(complete.cases(dat[, positions, drop = FALSE]))
  cat(sprintf("\nStarting %s\n  gene: %s; rows: %d; subtypes: %d; complete profiles: %d\n",
              basename(path), gene, nrow(dat), length(unique(dat$Subtype)), complete_profiles))
  cat(sprintf("  RAS positions: %s\n", paste(ras, collapse = ",")))

  data_sets <- list(all_sequences = dat, exact_deduplicated = exact_deduplicate(dat, positions))
  if (!is.null(params$near_duplicate_differences)) {
    label <- paste0("complete_linkage_", params$near_duplicate_differences, "diff")
    data_sets[[label]] <- complete_linkage_deduplicate(
      dat, positions, params$near_duplicate_differences
    )
  }
  summary <- data.frame(dataset = names(data_sets), n_rows = vapply(data_sets, nrow, integer(1)),
                        n_subtypes = vapply(data_sets, function(x) length(unique(x$Subtype)), integer(1)))
  prefix_base <- paste0(tools::file_path_sans_ext(basename(path)), "_covariation")

  for (data_name in names(data_sets)) {
    analysis_dat <- data_sets[[data_name]]
    cat(sprintf("  data set: %s (%d rows)\n", data_name, nrow(analysis_dat)))
    prefix <- paste(prefix_base, data_name, sep = "_")
    for (family_name in names(families)) {
      pairs <- families[[family_name]]
      cat(sprintf("  scan: %s (%d pairs)\n", family_name, nrow(pairs)))
      results <- analyze_pairs(analysis_dat, pairs, params$permutations, params$min_n,
                               params$min_minor, family_name)
      write_scan_outputs(prefix, family_name, results, analysis_dat)
    }
  }
  write.csv(summary, file.path(params$output_dir, paste0(prefix_base, "_dataset_summary.csv")), row.names = FALSE)
}
```

# Interpretation reminder

A small q value and positive bias-adjusted MI support an amino-acid association
within the sampled subtypes. They do not prove direct biological interaction:
shared ancestry, repeated sampling, and study or geographic clustering can
also create an association.
